---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##   I. Loading and preprocessing the data

1. Retrive the activity monitoring data sample from [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
,then load it into a dataframe called **dataRaw** using **read.csv**. 


```{r}{loadSample}

      dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      dataZip <- "./activity.zip"
      dataCsv <- "./activity.csv"

     if (!file.exists(dataCsv)){

           if(!file.exists(dataZip)){
               download.file(dataUrl, dataZip)
           }
           unzip(dataZip)
      }

      dataRaw <- read.csv(dataCsv)
      inpRows <- nrow(dataRaw)
      naRows  <- nrow(dataRaw[is.na(dataRaw$steps),])
      str(dataRaw)
      
```

2. Here is information/metadata about our activity dataset variables.

- **steps:**     Number of steps taking in a 5-minute interval (there are **`r naRows`** missing values are coded as **NA**)
- **date:**      The date on which the measurement was taken in *YYYY-MM-DD* format
- **interval:**  Identifier for each 5-minute interval in which a measurement was taken
- There are a total of **`r inpRows`** observations in this dataset.

3. In order to evalutate dates properly, we next cast **date** from *Factor* to *Date* storing our
modified dataframe in **sample1**. 

```{r}{processSample}

      sample1 <- dataRaw
      sample1$date <- as.Date(sample1$date)
      str(sample1)
```
- **sample1** will be used to answer the next two questions.

##  II. What is mean total number of steps taken per day?

Loading the dplyr library for processing the data frames

```{r} 
   library(dplyr)
```

1. To answer this question, we need to aggregate the *sum* (total) number of steps taken each **date** from the **sample1** dataframe. We also capture the **mean** and **median** number of steps taken per day. Our aggregated result set will be stored in **totStepsXDay**

```{r}{calc_totStepsXDay}

   totStepsXDay  <- sample1                %>%
                    group_by(date)         %>%
                    filter(!is.na(steps))  %>%
                    summarise(steps = sum(steps, na.rm=TRUE))
   
   meanSteps1    <- as.character(round(mean(totStepsXDay$steps,na.rm=T ),1))
   medianSteps1  <- as.character(round(median(totStepsXDay$steps,na.rm=T ),1))
   
   head(totStepsXDay)
   summary(totStepsXDay$steps)
   
```

2. To view distribution and visualize the mean total number of steps taken per day, we are presenting **totStepsXDay** in a histogram

```{r}{plotTotSteps}

   plot( totStepsXDay$date, 
         totStepsXDay$steps, 
         type="h", 
         main="Histogram of Steps per Day", 
         xlab="Date", 
         ylab="Steps", 
         #col="blue", 
         lwd=8)

  abline(h=mean(totStepsXDay$steps, na.rm=TRUE), col="red", lwd=1)

```

- The **mean** number of steps taken per day is **`r meanSteps1`** 
- The **median** number of steps taken per day is **`r medianSteps1`**

## III. What is the average daily activity pattern?

1. To answer this question, we need to aggregate the *mean* (average) number of steps taken each **interval** from the **sample1** dataframe. The aggregated data frame is stored in **avgStepsXInt** for additional examination.  We also capture Which 5-minute interval, on average, across all days in the **sample1**, contains the maximum number of steps.

```{r}{calcTotaleps}

   avgStepsXInt <- sample1                   %>%
                   group_by(interval)        %>%
                   filter(!is.na(steps))     %>%
                   summarise(steps = mean(steps, na.rm=TRUE))
   
   maxStepInt   <- avgStepsXInt$interval[which(avgStepsXInt$steps == max(avgStepsXInt$steps))]
   head(avgStepsXInt)
   summary(avgStepsXInt$steps)
     
```

2. The graph below is a time series plot of the 5-minute intervals (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r}{plotAvgSteps}

   plot( avgStepsXInt$interval, 
         avgStepsXInt$steps, 
         type="l", 
         main="Average Daily Steps Taken", 
         xlab="5-Minute Interval", 
         ylab="Avg Steps Taken", 
         col="blue", 
         lwd=1)

  abline(v=maxStepInt, col="red", lwd=0.5)
  
```

- The 5-minute interval,that contains the maximum number of steps is **`r maxStepInt`**

##  IV. Imputing missing values

1. We observed in **I. Loading and preprocessing the data** there are **`r naRows`** observations missing values for **steps** (coded as NA) in the **dataRaw** and **sample1**. The presence of missing data may introduce bias into our aggragations *(summaries/averages)*. To evaluate the impact of the **`r naRows`**  missing values, we substitute each missing value for number of steps taken for each day, with a calculated mean number of steps for each day. our modified dataframe inputing missing values will be dataframe **sample2**.

```{r}{ImputeMissingValues}

   sample2       <- sample1
   mvalues       <- is.na(sample2$steps)
   avg_interval  <- tapply( sample2$steps, 
                            sample2$interval, 
                            mean, 
                            na.rm=TRUE, 
                            simplify = TRUE )
   
   sample2$steps[mvalues] <- avg_interval[as.character(sample2$interval[mvalues])]
   
   mvalues <- nrow(sample2[is.na(sample2$steps),])
   
   head(sample2)
   str(sample2)

```
2. Using the same method (as in step **II. What is mean total number of steps taken per day** above) we will now aggregate the *sum* (total) number of steps taken each **date** from the **sample2** dataframe containing our imputed missing steps. We also capture the **mean** and **median** number of steps taken per day. Our aggregated result set will be stored in **totStepsXDay2**

```{r}{calc_totStepsXDay2}

   totStepsXDay2  <- sample2               %>%
                    group_by(date)         %>%
                    filter(!is.na(steps))  %>%
                    summarise(steps = sum(steps, na.rm=TRUE))
   
   meanSteps2    <- as.character(round(mean(totStepsXDay2$steps,na.rm=T ),1))
   medianSteps2  <- as.character(round(median(totStepsXDay2$steps,na.rm=T ),1))
   
   head(totStepsXDay2)
   summary(totStepsXDay2$steps)
   
```
3. To view distribution and visualize the mean total number of steps taken per day, we are presenting **totStepsXDay2** in a histogram

```{r}{plotTotSteps2}

   plot( totStepsXDay2$date, 
         totStepsXDay2$steps, 
         type="h", 
         main="Histogram of Steps per Day", 
         xlab="Date", 
         ylab="Steps", 
         col="blue", 
         lwd=8)

  abline(h=mean(totStepsXDay$steps, na.rm=TRUE), col="red", lwd=1)

```

- The **mean** number of steps taken per day is now **`r meanSteps2`** 
- The **median** number of steps taken per day is now **`r medianSteps2`**

*The impact of imputing missing data with the average number of steps in the same 5-min interval is that* 
*both the* **mean** *and the* **median** *are same value :* **`r meanSteps2`** *

##   V. Are there differences in activity patterns between weekdays and weekends?

1. First we will add a variable called **type** to **sample2** by deriving the whether a particular date falls on a **weekday**, or **weekend**.

```{r}{calcWeektype}

   sample2  <- sample2  %>%
               mutate(type = ifelse(weekdays(sample2$date)=="Monday"    | 
                                    weekdays(sample2$date)=="Tuesday"   |
                                    weekdays(sample2$date)=="Wednesday" |
                                    weekdays(sample2$date)=="Thursday"  |
                                    weekdays(sample2$date)=="Friday",
                                    "Weekday", 
                                    "Weekend"))

   head(sample2)
   str(sample2)
   
```


2. To answer this question, we will aggregate the *mean* (average) number of steps taken by each **interval** and **type** from the **sample2** dataframe. The aggregated data frame is stored in **avgStepsXInt** for additional examination.


```{r}{aggXIntXtype}

   avgStepsXIntXType <- sample2                   %>%
                        group_by(interval, type)  %>%
                        summarise(steps = mean(steps, na.rm=TRUE))
 

   head(avgStepsXIntXType)
   str(avgStepsXIntXType)
   
```

3. Finally we make a panel plot containing a time series plot **(i.e. type = "l")** of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}{plotavgStepsXIntXType}
    
    library(ggplot2)

    p <- ggplot( avgStepsXIntXType, 
                 aes( x     = interval, 
                      y     = steps))                                +                    
                 geom_line()                                         +
                 labs( title = "Avg. Daily Steps by Day of Week Type", 
                       x     = " 5 - Minute Interval", 
                       y     = "Avg No. of Steps Taken")             +
                  facet_wrap( ~type, 
                              ncol = 1, 
                              nrow = 2)
    print(p)
    
```

*The panel plot above illustrates there * **are** *differences in activity patterns between weekdays and weekends*

  
