---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##   I. Loading and preprocessing the data

The first step retrieves a data sample (if necessary) and loads it using read.csv for processing and analysis.

The 52K data sample for this analysis is downloaded from the web site: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

```{r}{loadSample}

      dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      dataZip <- "./activity.zip"
      dataCsv <- "./activity.csv"

     if (!file.exists(dataCsv)){

           if(!file.exists(dataZip)){
               download.file(dataUrl, dataZip)
           }
           unzip(dataZip)
      }

      sample <- read.csv(dataCsv)
```

Variables included in the source dataset are:

1. **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. **date:** The date on which the measurement was taken in YYYY-MM-DD format
3. **interval:** Identifier for the 5-minute interval in which measurement was taken

There should be a total of **17,568** observations in this dataset.

```{r}{examineSample}
      str(sample)
```

##  II. What is mean total number of steps taken per day?



## III. What is the average daily activity pattern?



##  IV. Imputing missing values



##   V. Are there differences in activity patterns between weekdays and weekends?
