---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##   I. Loading and preprocessing the data

The first step retrieves a data sample (if necessary) and loads it using read.csv for processing and analysis.  The 52K data sample for this analysis is downloaded from the web site: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

```{r}{loadSample}

      dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      dataZip <- "./activity.zip"
      dataCsv <- "./activity.csv"

     if (!file.exists(dataCsv)){

           if(!file.exists(dataZip)){
               download.file(dataUrl, dataZip)
           }
           unzip(dataZip)
      }

      sample <- read.csv(dataCsv)
```

Variables included in the source dataset are:

1. **steps:**     Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. **date:**      The date on which the measurement was taken in *YYYY-MM-DD* format
3. **interval:**  Identifier for the 5-minute interval in which measurement was taken

There should be a total of **17,568** observations in this dataset.

```{r}{examineSample}
      str(sample)
```

This next step processes the data for the analysis below by converting **date** from *factor* to *date*

```{r}{processSample}
      sample$date <- as.Date(sample$date)
      str(sample)
```


##  II. What is mean total number of steps taken per day?

To answer this question, we need to aggregate (*sum*) the total number of steps taken each day. We will also calculate the **mean** and **median** number of the total number of steps taken per day


```{r}{calcTotalSteps}

   library(dplyr)
   TotStepsXDay  <- sample                %>%
                    group_by(date)        %>%
                    filter(!is.na(steps)) %>%
                    summarise(steps = sum(steps, na.rm=TRUE))
   
   meanSteps    <- as.character(round(mean(TotStepsXDay$steps,na.rm=T ),1))
   medianSteps  <- as.character(round(median(TotStepsXDay$steps,na.rm=T ),1))
   
   head(TotStepsXDay)
   summary(TotStepsXDay$steps)
   
     
```


Below are the total steps per day presented in a histogram to see the distribution of steps over days, and visualize 
the mean.


```{r}{plotTotSteps}

   plot( TotStepsXDay$date, 
         TotStepsXDay$steps, 
         type="h", 
         main="Histogram of Steps per Day", 
         xlab="Date", 
         ylab="Steps", 
         #col="blue", 
         lwd=8)

  abline(h=mean(TotStepsXDay$steps, na.rm=TRUE), col="red", lwd=1)

```

- The **mean** number of steps taken per day is **`r meanSteps`** 
- The **median** number of steps taken per day is **`r medianSteps`**


## III. What is the average daily activity pattern?

To answer this question, we need to aggregate the (*mean or average*) number of steps taken each **interval**. We will also calculate which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps

```{r}{calcTotaleps}

   AvgStepsXInt <- sample                    %>%
                   group_by(interval)        %>%
                   filter(!is.na(steps))     %>%
                   summarise(steps = mean(steps, na.rm=TRUE))
   
   MaxSteps    <- AvgStepsXInt$interval[which(AvgStepsXInt$steps == max(AvgStepsXInt$steps))]
   
   head(AvgStepsXInt)
   summary(AvgStepsXInt$steps)
     
```

The graph below plots the average steps taken per time interval across days.  

```{r}{plotAvgSteps}

   plot( AvgStepsXInt$interval, 
         AvgStepsXInt$steps, 
         type="l", 
         main="Average Daily Steps", 
         xlab="Interval", 
         ylab="Steps", 
         col="blue", 
         lwd=1)

  abline(v=MaxSteps, col="red", lwd=0.5)
  
```

- The 5-minute interval,that contains the maximum number of steps is **`r MaxSteps`**

##  IV. Imputing missing values

```{r}{calMV}
   missingVal <- nrow(sample[is.na(sample$steps),])
```

We noticed there are a number of days/intervals (**`r missingVal`**) that are missing values for steps (coded as **NA**). The presence of missing data may introduce bias into some calculations or summaries of the data.  Impute missing values using mean for each day.

```{r}{ImputeMissingValues}

   imputedSample <- sample
   missingValues <- is.na(imputedSample$steps)
   avg_interval  <- tapply( imputedSample$steps, 
                            imputedSample$interval, 
                            mean, 
                            na.rm=TRUE, 
                            simplify = TRUE )
   
   imputedSample$steps[missingValues] <- avg_interval[as.character(imputedSample$interval[missingValues])]
   
   missingVal <- nrow(imputedSample[is.na(imputedSample$steps),])
   
   head(imputedSample)
   str(imputedSample)

```

In our new dataset there are now (**`r missingVal`**) days/intervals missing values for steps (coded as **NA**). 

Below are the total steps per day presented in a histogram to see the distribution of steps over days, and visualize 
the mean.

```{r}{calcTotalImpSteps}

   TotStepsXIDay  <- imputedSample         %>%
                     group_by(date)        %>%
                     summarise(steps = sum(steps, na.rm=TRUE))
   
   meanImpSteps    <- as.character(round(mean(TotStepsXIDay$steps,na.rm=T ),1))
   medianImpSteps  <- as.character(round(median(TotStepsXIDay$steps,na.rm=T ),1))
   
   head(TotStepsXDay)
   summary(TotStepsXDay$steps)
   
     
```

```{r}{plotImpSteps}

   plot( TotStepsXDay$date, 
         TotStepsXDay$steps, 
         type="h", 
         main="Histogram of Steps per Day", 
         xlab="Date", 
         ylab="Steps", 
         #col="blue", 
         lwd=8)

  abline(h=mean(TotStepsXDay$steps, na.rm=TRUE), col="red", lwd=1)

```

- The **mean** number of steps taken per day is **`r meanImpSteps`** 
- The **median** number of steps taken per day is **`r medianImpSteps`**

What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}{calcImpact}

 diffMean    <- as.character(round(mean(TotStepsXIDay$steps,na.rm=T ),1) 
                           - round(mean(TotStepsXDay$steps,na.rm=T ),1))

 diffMedian  <- as.character(round(median(TotStepsXIDay$steps,na.rm=T ),1) 
                           - round(median(TotStepsXDay$steps,na.rm=T ),1))

```

- The difference in **mean** number of steps between the modified and raw sample is **`r diffMean`**
- The difference in **median** number of steps between the modified and raw sample is **`r diffMedian`**

The impact of imputing missing data with the average number of steps in the same 5-min interval is that 
both the mean and the median are same value : **`r meanImpSteps`** 



##   V. Are there differences in activity patterns between weekdays and weekends?
